---
title: "Toronto marriage data"
subtitle: ""
author: 
  - David Qi
thanks: "Code and data are available at: https://github.com/UTDQi/reflection5."
date: today
date-format: long
abstract: "Practice document for week three. Includes a brief explaination of the toronto marriage dataset and a graph."
format: pdf
number-sections: true
bibliography: references.bib
---


```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(spotifyr)

the_national <- readRDS("data/raw/the_national.rds")

Elvis_Presley = readRDS("data/raw/Elvis_Presley.rds")

Adele = readRDS("data/raw/Adele.rds")
```

# Introduction

In this paper, we make use of R (@citeR) and the tidyverse package (@tidyverse) with supporting package dplyr(@dplyr) to analysis data form the 2022 ACS(American Community Surveys). The data set is from spotify API (@spotify_api), gathered using spotifyr (@spotifyr).  We have used some information and codes form the book Telling Stories With Data(@tellingstories) on accessing API. 

The remainder of this paper is structured as follows: We talk about how the data is gained under [Data](#sec-data) section. And describe the method of data estimation we use in the [Method](#method) section. Then, we provide a table that compares our estimation with actual data in the [Results](#results) section. We then discusses reason of error in estimation [Discussion](#discussion) section. 


# Data {#sec-data}

Some of our data (@fig-bills) are from @opendatatoronto.

```{r}
#| label: fig-bills
#| fig-cap: Month of marriage
#| echo: false

ggplot(table_month, aes(x = month, y = total_marriage)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Number of Marriages per Month",
       x = "Month",
       y = "Number of Marriages") +
  scale_x_continuous(breaks = 1:12)+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels
```



\newpage

# References


