---
title: "Analysis on Elvis Presley: Increased Energy Level of The Fans"
subtitle: ""
author: 
  - David Qi
  - Chendong Fei
thanks: "Code and data are available at: https://github.com/UTDQi/Elvis_Analysis"
date: today
date-format: long
abstract: "This paper analysised the albums of Elvis Presley on spotify. We find that as the energy of these albums have increased over the years, similar trend is not seen on tempo. Which indicates that as taste of music of Elvis fans have shifted toward more energetic. The favor of tempo have not changed."
format: pdf
number-sections: true
bibliography: references.bib
---


```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(spotifyr)
library(gridExtra) 

Elvis_Presley <- readRDS(here::here("data/raw/Elvis_Presley.rds"))
```

# Introduction

Elvis Presley (1935-1977) is an American celebrity who gained immense popularity worldwide beginning in the 1950s. His enduring fame allows us to analyze how audience tastes have changed over the years. We find that the energy of these albums has increased over time, while tempo has remained constant. This suggests that Elvis fans' musical preferences have shifted toward more energetic music, while their preference for tempo has not changed.

In this paper, we make use of R [@citeR] and the tidyverse package [@tidyverse], with support from the dplyr package [@dplyr], to analyze Spotify (an online music platform launched in 2008) data related to Elvis Presley (1935-1977). The dataset was obtained from the Spotify API [@spotify_api] using the spotifyr package [@spotifyr]. The graphs were created using the ggplot2 [@ggplot] and gridExtra [@gridExtra] packages in R. We have also used some information and code from the book Telling Stories With Data [@tellingstories] for accessing the API.

The remaining part of this paper is structured as follows: We discuss the basic aspects of the data in the [Data](#sec-data) section.  Then, we analysis the data in the [Results](#sec-results) section. We then discusses our finding and draw on possible limitations in the [Conclusions](#sec-conclusion) section. 


# Data {#sec-data}
In this section, we discuss aspects of the data and outline how it will be used.

As shown in [@fig-albums], the release dates of albums span from 1956 to 2024. Note that album releases were suspended in 1977 (marked by the red dotted line) when Elvis Presley passed away, and resumed after 2010. This resumption is due to Spotify's new licensing agreements, which allowed the release of new albums by classic artists.

There were also many albums released in 1997, 20 years after Elvis Presley passed away. These albums were released to recognize this event.

```{r}
#| label: fig-albums
#| tbl-cap: Albums Released Each Year
#| include: true
#| echo: false
#| warning: false
#| message: false

year = Elvis_Presley %>%
  mutate(release_year = as.numeric(strtrim(album_release_date,4)))

albums_per_year <- year %>%
  group_by(release_year) %>%
  summarise(album_count = n())

albums_per_year |>
  ggplot(aes(x = release_year, y = album_count)) +
  geom_col(alpha = 0.7) +
  geom_vline(xintercept = 1977, linetype = "dashed", color = "red", size = 1) +
  theme_minimal()+
  labs(
    x = "Album release date",
    y = "Valence"
  ) +
  theme(legend.position = "bottom")
```

In this paper, we will analyze these two groups of albums to examine how people's tastes have changed over the past 30 years. We note that albums released during this time reflect the general preferences of the audience, as album producers tend to select and remix songs based on prevailing tastes.

To represent the abstract concept of 'taste,' we will use data on the energy and tempo of the music. It’s important to note that this data is generated by a machine learning model about which we have limited information on. As a result, it may not fully capture the actual 'energy' or 'tempo' of the music. However, it provides a useful generalization of these factors. The fact that Spotify uses this data indicates its usefulness for analysis.

# Results {#sec-results}

## Energy
We first analysis the energy. 
[@fig-energy] and [@fig-energy2] shows the change in energy across the years. 

```{r}
#| label: fig-energy
#| tbl-cap: Change in energy acrocess the years
#| include: true
#| echo: false
#| warning: false
#| message: false

# Plot the relationship between two variables
ggplot(Elvis_Presley, aes(x = album_release_year, y = energy)) +
  geom_point() +   
  geom_smooth() +# Scatter plot points
  labs(
       x = "Year Album released", 
       y = "Energy") +         # Axis labels and title
  theme_minimal()                  # Use a minimal theme for the plot

```
```{r}
#| label: fig-energy2
#| tbl-cap: Compaing Energy of The Two Time Periods
#| include: true
#| echo: false
#| warning: false
#| message: false
# Splitting the data into two parts based on a year (e.g., split before and after the year 2000)
data_part1 <- Elvis_Presley %>% filter(album_release_year < 2000)
data_part2 <- Elvis_Presley %>% filter(album_release_year >= 2000)

# Calculating the average energy for both parts
avg_energy_part1 <- mean(data_part1$energy, na.rm = TRUE)
avg_energy_part2 <- mean(data_part2$energy, na.rm = TRUE)

# Create histograms for the two parts and add mean lines
hist_part1 <- ggplot(data_part1, aes(x = energy)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = avg_energy_part1), color = "red", linetype = "dashed", size = 1) +  # Mean line for part1
  labs(title = "Energy Distribution (Before 2010)", x = "Energy", y = "Count") +
  theme_minimal() +
  annotate("text", x = avg_energy_part1, y = Inf, label = paste("Mean =", round(avg_energy_part1, 2)), vjust = -0.5, hjust = 1, color = "red")  # Annotate mean

hist_part2 <- ggplot(data_part2, aes(x = energy)) +
  geom_histogram(binwidth = 0.05, fill = "green", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = avg_energy_part2), color = "red", linetype = "dashed", size = 1) +  # Mean line for part2
  labs(title = "Energy Distribution (2010 and After)", x = "Energy", y = "Count") +
  theme_minimal() +
  annotate("text", x = avg_energy_part2, y = Inf, label = paste("Mean =", round(avg_energy_part2, 2)), vjust = -0.5, hjust = 1, color = "red")  # Annotate mean

# Arrange the two histograms side by side
grid.arrange(hist_part1, hist_part2, ncol = 2)


```
Initially, in the pre-1970s era, the energy levels of albums are relatively moderate, with a gradual increase reaching a peak around the 1980s. This trend continues after the year 2010, when the energy levels of albums rises again, reflecting a general increase in the intensity and dynamism of music in more recent years. For instance, as shown in [@fig-energy2], the mean energy level (represented by the red dashed lines) before 2010 is 0.467, while the mean energy level after 2010 has increased to 0.549. Overall, the graph suggests that while music energy has increased over time, more recent albums tend to have higher energy on average.

## Tempo
We now turn to analysis the tempo. 
[@fig-tempo] and [@fig-tempo2] shows the change in tempo across the years. 

```{r}
#| label: fig-tempo
#| tbl-cap: Change in tempo acrocess the years
#| include: true
#| echo: false
#| warning: false
#| message: false



# Plot the relationship between two variables
ggplot(Elvis_Presley, aes(x = album_release_year, y = tempo)) +
  geom_point() +   
  geom_smooth() +# Scatter plot points
  labs(x = "Year Album released", 
       y = "Tempo") +         # Axis labels and title
  theme_minimal()                  # Use a minimal theme for the plot

```
```{r}
#| label: fig-tempo2
#| tbl-cap: Compaing Tempo of The Two Time Periods
#| include: true
#| echo: false
#| warning: false
#| message: false
# Splitting the data into two parts based on a year (e.g., split before and after the year 2010)
data_part1 <- Elvis_Presley %>% filter(album_release_year < 2000)
data_part2 <- Elvis_Presley %>% filter(album_release_year >= 2000)

# Calculating the average tempo for both parts
avg_tempo_part1 <- mean(data_part1$tempo, na.rm = TRUE)
avg_tempo_part2 <- mean(data_part2$tempo, na.rm = TRUE)

# Create histograms for the two parts and add mean lines
hist_part1 <- ggplot(data_part1, aes(x = tempo)) +
  geom_histogram(binwidth = 5, fill = "blue", alpha = 0.7) +
  geom_vline(aes(xintercept = avg_tempo_part1), color = "red", linetype = "dashed", size = 1) +  # Mean line for part1
  labs(title = "Tempo Distribution (Before 2010)", x = "Tempo", y = "Count") +
  theme_minimal() +
  annotate("text", x = avg_tempo_part1, y = Inf, label = paste("Mean =", round(avg_tempo_part1, 2)), vjust = -0.5, hjust = 1, color = "red")  # Annotate mean

hist_part2 <- ggplot(data_part2, aes(x = tempo)) +
  geom_histogram(binwidth = 5, fill = "green", alpha = 0.7) +
  geom_vline(aes(xintercept = avg_tempo_part2), color = "red", linetype = "dashed", size = 1) +  # Mean line for part2
  labs(title = "Tempo Distribution (2010 and After)", x = "Tempo", y = "Count") +
  theme_minimal() +
  annotate("text", x = avg_tempo_part2, y = Inf, label = paste("Mean =", round(avg_tempo_part2, 2)), vjust = -0.5, hjust = 1, color = "red")  # Annotate mean

# Arrange the two histograms side by side
grid.arrange(hist_part1, hist_part2, ncol = 2)
```
For tempo, we see a different trend, the tempo of the musics selected to be included in the album of Elvis Presley have not changed much over the times. From [@fig-tempo2], we can see the mean of the tempo(the red dashed lines)  before 2010 is 112.78, and the mean after 2010 have have changed 112.94. Which is a very small change.The shape of the two graphs in [@fig-tempo2] are also similar, suggesting the selection of music have experienced no major change if we evaluate form tempo. [@fig-tempo] also showed the Smoothed conditional means close to a straight line, indicating that the tempo have not changed over time

# Conclusions {#sec-conclusion}

Form our [Results](#results) section, we see that the energy of the music have experienced a increase over the years. This signifies that, in the $21^{st}$ century, Elvis Presley songs that are more energetic tend to be selected into new albums, and people tend to make new remixes of these songs that are more energetic.

However, energetic does not mean fast. People's taste of tempo have not changed much over the time. And there is no similar favor over faster music. 

## Limitations

We note that Energy is a variable that is difficult to find the meaning to, and may not perfectly match our instinctive definition of "energy of music". Further information of the data is required to understand what "energy" accurately represent. Future studies should focus on aspects of the data that are more explainable and understandable.

Also, this analysis only reflect the change in musical tasted of Elvis Fans, which is very small proportion of the general population, future research may use data on more artists to gain a comprehensive on the subject. 

\newpage

# References


